---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lessLM

The sole purpose of this package is to demonstrate how the optimizers implemented in lessSEM can be used by other packages. To this end, we use a fairly simple model: A
linear regression of the form
$$\pmb y = \pmb X \pmb b + \pmb\varepsilon$$

## Step 1

Install lessSEM from https://github.com/jhorzek/lessSEM.

## Step 2

Create a new R package which uses RcppArmadillo (e.g., with ``RcppArmadillo::RcppArmadillo.package.skeleton()''). Open the DESCRIPTION file and add lessSEM to the "LinkingTo" field (see the DESCRIPTION file of this package if you are unsure what we are referring to). If you already have a package, just add lessSEM to the "LinkingTo" field.

## Step 3: Implementing your package

We will assume that you already have a package which implements the model that you would like to regularize. Therefore, we will not make use of any of the functions and classes defined in lessSEM at this point. However, make sure that you have two functions:

1) A function which computes the fit (e.g., -2-log-Likelihood) of your model. This function must return a double.
2) A function which computes the gradients of your model. This function must return an arma::rowvec.

For our example, we have implemented these functions in the files src/linearRegressionModel.h and src/linearRegressionModel.cpp. We also implemented a function to compute the Hessian. This function is very helpful when using the glmnet optimizer which does use an approximation of the Hessian based on the BFGS procedure. If the initial Hessian is a poor substitute for the true Hessian, this optimizer can return wrong parameter estimates. To get a reasonable initial Hessian, we therefore also implemented this initial Hessian estimation based on the procedure used in lavaan.

## Step 4: Linking to lessSEM

Assuming that our model is set up, we are ready to link everything to lessSEM. First, create an new file (we called ours src/optimization.cpp). Here, it is important to include the lessSEM.hpp headers (see src/optimization.cpp). All further steps necessary to use the lessSEM optimizers are outlined in the comments included in the new file src/optimization.cpp. Open this file and follow the instructions.

## Step 5: Test your function

```{r}
set.seed(123)
library(lessLM)

# first, we simulate data for our
# linear regression.
N <- 100 # number of persons
p <- 10 # number of predictors
X <- matrix(rnorm(N*p),	nrow = N, ncol = p) # design matrix
b <- c(rep(1,4), 
       rep(0,6)) # true regression weights
y <- X%*%matrix(b,ncol = 1) + rnorm(N,0,.2)

l1 <- elasticNet(y = y,
                  X = X,
                  alpha = 1, 
                  lambda = 1)
l1

library(glmnet)
l2 <- glmnet(x = X, 
             y = y, 
             lambda = 1,
             standardize = FALSE)
coef(l2)[,1]
```